A Mountaintop View Requires Minimal Sorting: \break A Faster Contour Tree Algorithm \footnote{The full updated version of this paper is available online \url{\si{http://web.engr.illinois.edu/\string~\si{raichel2}/\si{paint.pdf}}}}

### 2015-03-31 paint-contour-tree.tex
small change

Our algorithm requires numerous ideas: partitioning the contour tree into join and split trees, 
a local growing procedure to iteratively build contour trees, and the use of heavy path 
decompositions for the time complexity analysis. There is a crucial use of a family of binomial heaps to 
maintain priorities, ensuring that any comparison made is between nodes in the same descending path in the contour tree.
We also prove lower bounds showing that the $\sum_{p\in P} |p|\log|p|$ complexity
is inherent to computing contour trees. 
%Consider the family of instances where our algorithm
%requires $C$ comparisons. Any algorithm computing contour trees on this family requires
%$\Omega(C)$ comparisons in the worst case.

%
%general lower bound of
%$\Omega(\sum_{p \in P} |p|\log |p|)$ for computing contour trees, showing
%that the tree structure must play a role in the time complexity.

\ignore{
Moreover, we show there   
exists some partition $P'$ such that the running time of our algorithm is 
$O(\sum_{p\in P'} |p|\log|p| + N\alpha(N))$, where $N$ is the total complexity of the input manifold

All previous contour tree algorithms involve a global sort 
of the heights of all points or critical vertices. Thereby, these algorithms always pay $\Omega(t \log t)$.
This is a major roadblock for efficiency, since sorting is the most expensive operation

Reeb graphs are a fundamental topological structure which tracks the evolution of levels sets of a real valued function over a manifold.  
In the case when this graph is loop free it is called a contour tree.  Due to the use of the contour trees in practice on large data sets, 
there has been significant previous work in developing algorithms for their efficient computation.

Here we present the first contour tree algorithm which takes into account the shape of the contour tree, 
and in doing so yields instance optimal results.
Specifically, we prove that for any partition of the contour tree into a set of descending paths, $P$, 
any comparison based algorithm requires $\sum_{p\in P} |p|\log|p|$ time.  Moreover, we show there   
exists some partition $P'$ such that the running time of our algorithm is 
$O(\sum_{p\in P'} |p|\log|p| + N\alpha(N))$, where $N$ is the total complexity of the input manifold.
In particular, our algorithm runs in $O(N\alpha(N))$ time when the tree is balanced, 
while previous algorithms ran in $O(N+t\log t)$ (where $t$ is the size of the contour tree).
%In particular, our algorithm runs in $O(N\alpha(N))$ time when the tree is balanced, yet still 
%matches the previous worst case $O(N+n\log n)$ bound when the tree is long and skinny.

In order to produce our instance optimal results we introduce a novel linear time surface partitioning scheme which is likely to have applications elsewhere.
Moreover, in order to prove the optimality we use a non-trivial combinatorial discharging argument which we believe is of independent interest and in particular uses 
such tools as heavy-light path decompositions.
}
\end{abstract}

@@

In a contour tree $T$, a \emph{leaf path} is simply a contiguous path in $T$
containing a leaf, which is also monotonic in the height values of its vertices. 
%A leaf path runs either upwards or downwards, never both.
%
% Then in the following two subsections 
%we relate path decomposition costs (defined below) to the running time of our algorithm using $\Lem{algCost}$.

@@

We complement \Thm{main-alg} with a lower bound, suggesting that the comparison complexity
of $\sum_{p \in P(T)} |p|\log |p|$ is inherent for any contour tree algorithm. The exact phrasing
of the lower bound requires some care. Consider the $d=2$ (terrain) case.
%
%We define a class of valid path decompositions.  This include any decomposition of rooted trees,
%and more generally is a large superset of all decomposition produced by our 
%algorithm.
%\footnote{Our construction likely works for any decomposition, however, 
%to keep the presentation simple we restrict to a subset relevant for our algorithm.}
%
Let a path decomposition $P$ be called \emph{valid} if it is actually output by the algorithm
on some input.
%
%
%As referenced in \Thm{main-alg}, for any $f:\MM \mapsto \RR$ our deterministic algorithm defines a 
%specific path decomposition $P(T)$%
\ignore{\FullVer{
\footnote{The specific decomposition used is defined in \Sec{pathDecomp}, 
but also depends on the raining and painting procedures described in \Sec{rain} and \Sec{paint}.  
The important thing is that our algorithm is deterministic, and so for a fixed $\MM$ and $f$ it is well defined.}.}
\SubmitVer{
\footnote{The specific decomposition used is defined in the full version of the paper.  
The important thing is that our algorithm is deterministic, and so for a fixed $\MM$ and $f$ it is well defined.}.}
We define $\cost(f) = \sum_{p \in P(T)} |p|\log |p|$ for this specific decomposition $P(T)$.
Ideally, we would like to argue that any contour-tree algorithm
must pay $\cost(f)$ for ant $f$. 
But this is obviously false, since one could consider an algorithm
that hard codes the answer for $f$ and merely checks this answer.

Instead, call a possible cost $C$ \emph{valid} if $C = \cost(f)$ for some $f$\footnote{If one is concerned
that $C$ might be irrational, define valid if $2^C = 2^{\cost(f)}$.}. We prove
a refined lower bound matching the performance of \Thm{main-alg}. 
Roughly speaking, it is not possible to improve on \Thm{main-alg} even when restricted to a subset
of inputs of fixed cost.

%\begin{definition}\label{def:hfdt}
%A \emph{height function decision tree} is an algorithm
%that is allowed arbitrary computation on the domain $\MM$, and is an algebraic decision
%tree over the heights. The running time is simply the number of queries made.
%\end{definition}
%
}
\begin{theorem} 
\label{thm:main-lb}
%Fix a sufficiently large numbers $t$ and $C$, 
%where $t$ is the number of critical points and $C$ is a valid cost for an instance with $t$ critical point.
Consider a valid path decomposition $P$. There exists a family $\bF_P$ of terrains ($d=2$) with the following properties.
Any contour tree algorithm makes $\Omega(\sum_{p \in P} |p|\log |p|)$ comparisons in the worst case over $\bF_P$.
Furthermore, for any terrain in $\bF_P$, the algorithm of \Thm{main-alg} makes $O(\sum_{p \in P} |p|\log |p|)$ comparisons.
%
%the set of functions $\bF_C$ that have the same number of critical points and the same (valid) cost $C$.  Any algebraic decision tree that correctly computes
%the contour tree on all of $\bF_C$ has a worst case running time of $\Omega(C)$.
\end{theorem}

@@

% We complement \Thm{main-alg} with an almost matching lower bound, showing that the complexity
% of $\sum_{p \in P(T)} |p|\log |p|$ is required for any contour tree algorithm. The exact phrasing
% of the lower bound requires some care. Consider the $d=2$ (terrain) case, and fix a mesh $\MM$.
% Now any height function $f$ (together with $\MM$) defines a specific contour tree $T$.
% As referenced in \Thm{main-alg}, given $f$, our algorithm then defines some specific path 
% decomposition $\overline{P}(T)$, such that the running time of our algorithm is bounded by 
% $O(\cost(\overline{P}(T))+t\alpha(t)+N)$, where $\cost(P(T)) = \sum_{p \in P(T)} |p|\log |p|$.
% A priori, which path decomposition this will be is unclear\footnote{The specific decomposition, defined in \Sec{pathDecomp}, 
% depends on arbitrary choices made during the raining and painting procedures described in \Sec{rain} and \Sec{paint}.}, 
% however we know there is some large set of potential path decompositions it lies in
% (the exact set is defined in \Def{path2}, but for now one can think of the set of all decompositions).
% Let $\costs(f)$ denote the set of costs of all these potential path decompositions that $f$ defines
% (i.e. $\cost(\overline{P}(T))\in \costs(f)$).
% 
% Ideally, we would like to argue that any contour-tree algorithm must pay $\cost(\overline{P}(T))$ for $f$. 
% But this is obviously false, since one could consider an algorithm
% that hard codes the answer for $f$ and merely checks this answer.
% Instead, call a possible cost $C$ \emph{valid} if $C \in \costs(f)$ for some $f$ 
% defined over the fixed mesh $\MM$\footnote{If one is concerned
% that $C$ might be irrational, define valid if $2^C = 2^{\cost(f)}$.}. We prove
% a refined lower bound matching \Thm{main-alg}. 
% 
% \begin{definition} \label{def:hfdt}
% A \emph{height function decision tree} is an algorithm
% that is allowed arbitrary computation on the domain $\MM$, and is an algebraic decision
% tree over the heights. The running time is simply the number of queries made.
% \end{definition}
% 
% \begin{theorem} \label{thm:main-lb} Consider the set of functions $\bF_C$ (defined over a fixed mesh $\MM$) 
% that all have the same value $C$ as a valid cost.  Any height function decision tree that correctly computes
% the contour tree on all of $\bF_C$ has an average case (uniform over $\bF_C$) running time of $\Omega(C)$.
% \end{theorem}

@@

\ignore{
classic output sensitive convex hull algorithms of Kirkpatrick and Seidel~\cite{KiSe86} and Chan~\cite{c-ooscha-96}.

\paragraph{Output Sensitive Analysis.}
In geometric settings often the goal is to compute a (potentially complex) geometric structure determined 
by the input, and so it is natural to also consider how the form of this output structure affects the running time.
One example is the standard sweep line algorithm of Bentley and Ottmann \cite{bo-arcgi-79} for line segment intersection, 
which runs $O((n+k)\log n)$ time where k is the number of intersections, which in the worst case is quadratic.
Another example is Chan's algorithm for computing the convex hull \cite{c-ooscha-96} which runs in $O(n\log h)$ 
where $h$ is the number of vertices on the hull, which in the worst case is $n$.  While such algorithms represent 
significant improvements over their worst case counterparts, their running times are only measured with respect 
to the size of the output and not any structure it may have.

\paragraph{Instance Optimality.}  
Here we are concerned with a more refined analysis than the typical output sensitive analysis.
Specifically, we investigate an algorithm which is instance optimal as defined by Afshani \etal \cite{abc-ioga-09}.  
In order to prove such a result, one considers a family of allowable partitions of the input, 
each of which is defined in terms of the structure of the output.  Each such partition then has 
an associated entropy type cost representing the number of possible instantiations of such a partition.
Our work differs slightly from Afshani \etal \cite{abc-ioga-09} however, in that they show any algorithm 
(for say the convex hull) requires the minimum such cost, where as in our case it is shown that 
any algorithm requires the maximum over these costs.
}

@@

This is convenient, as we can now assume that $f$ is defined on $\RR^d$. Any point inside a boundary facet
has a well-defined height, and this includes the infinite facet. 
We think of $d$ as constant, and assume that $\MM$ is represented in a data structure that allow constant-time traversals of neighboring simplices
in $\MM$ (e.g.~\cite{BoMa12}). (This is analogous to a DCEL for higher dimensions.)
Observe that $f:\MM \rightarrow \RR$ can be thought of as a $d$-dimensional simplicial complex living in $\RR^{d+1}$.
We can think of $f(x)$ as the ``height" of a point $x \in \MM$, and encoded in the representation of $\MM$. 
%Abusing notation, $\MM$ refers to this encoding, so we do not refer to $f$.

@@

% \begin{definition} \label{def:deg} The up-degree of $x \in \MM$ is the number of $(f(x) + \delta)$-contours that intersect $B_\eps(x)$
% as $\delta, \eps \rightarrow 0^+$. The down-degree is the number of $(f(x) - \delta)$-contours that intersect $B_\eps(x)$
% as $\delta, \eps \rightarrow 0^+$.
% \end{definition}

@@

We use $\cV(f)$ to denote the set of critical points.
Because $f$ is piecewise-linear, all critical points are vertices in $\MM$. 
A value $h$ is called critical, if $f(v) = h$, for some $v \in \cV(f)$. 
The critical points are exactly where the topology of level sets change.
By assuming that our manifold is boundary critical, the vertices on a given boundary are either collectively all maxima or all minima.  
We abuse notation and refer to this entire set of vertices as a maxima or minima.
%In particular, we do not view this as a violation of the Morse condition on $f$, which requires distinct heights for critical points.

\ignore{
The Morse assumption
implies that all critical points have distinct values. A contour that involves a critical
point is also denoted as critical. Note that the vertices in a boundary face collectively behaves like an extremum.
Abusing notation, the term ``critical point", ``maxima", and ``minima" will include boundaries.
Critical values include these boundary values. By boundary criticality, all contours are themselves
represented as $d-1$-dimensional simplicial complexes.
}

@@

\ignore{
\begin{remark}
 Note that \Def{deg} can be changed so that the up-degree (resp. down-degree) is the number of $(f(x) + \delta)$-contours (resp. $(f(x) - \delta)$-contours) that intersect $B_\eps(x)$.
 For $2$-manifolds, the two definitions are equivalent, and at each saddle point either two contours join together or split apart, and hence the terminology.  
 In higher dimensions though these definitions differ.  Specifically, this alternative definition only includes points at which the number of contours in a level set change.
 These points are a subset of our defined critical points which also include points at which the topology of the contours change (and hence the term `critical' is still applicable).
 Note that our critical point definition allows one to classify all vertices (according to \Def{points}) simultaneously in linear time, as it is a local neighborhood condition.
\end{remark}

\begin{remark}
Note that the contour tree as defined above contains vertices for all critical points, 
which differs slightly from the usual definition which only contains critical points corresponding to a change in the number of contours.  Specifically, if a vertex in the tree we defined has degree $2$ then only the topology of a contour is changing rather than the number of contour.  As such the standard contour tree can be computed from ours by a linear time pass to contract maximal degree 2 vertex paths.

For simplicity we will also make the standard assumption that vertices in the contour 
tree have degree at most $3$, i.e. there are no multi-saddles.
\end{remark}
}

@@

%
%{\bf Punchline:} We give a linear time procedure that partitions $\MM$ into smaller pieces. We guarantee
%that the contour tree of each individual piece is basically the join (or split) tree. This saves
%some of the redundant computation. We then show to put it all together the get the full contour tree.
%Now for more details.

@@

An \emph{initial} painting has the additional property that the restriction $\chi:X \mapsto [|X|]$ is a bijection.
\end{definition}
%
%Paintings are not unique. It is straightforward to construct an initial painting of $\MM$ in $O(|\MM|)$ time, 
%using a descending BFS from each maximum that does not explore previously colored edges. 
%As the algorithm proceeds, the bijection property will not hold, though we always maintain a valid painting.

@@

\ignore{
\medskip
\fbox{
\begin{minipage}{0.9\textwidth}
{\bf $\cut(\phi,\MM)$}

\smallskip
\begin{asparaenum}
%	\item Insert the vertices of $\phi$ in $\MM$.
	\item For every edge $e$ that intersects $\phi$ at point $x$: add edges $(e,x^+)$ and $(e,x^-)$
	where $x^+$ (resp. $x^-$) is infinitesimally above (resp. below) $x$.
	\item Take a point $x$ (resp. $y$) on $e$ at distance $\eps$ above (resp. below) $\phi$.
	\item Let $\phi_x$ be the contour through $x$, and analogously define $\phi_y$. Insert a vertex
	at each intersection point of these contours with edges of $\MM$. Add  
	\item Delete the edges between $\phi_x$ and $\phi_y$.
\end{asparaenum}
\end{minipage}}

\medskip
This results in a new boundary critical simplicial complex. There are two new disjoint
boundary simplices added. The manifold may be disconnected. 
}

@@

\ignore{
\begin{proof} If $y \in \wet(x,\MM)$, all the points in any contour containing $y$ are also in $\wet(x,\MM)$.
 (Follow the non-ascending path to from $x$ to $y$ and then walk along the contour.)
 First observe that any point which in the interface must be wet.  Second, for $y \in \wet(x,\MM)$, all the points in any contour containing $y$ are also in $\wet(x,\MM)$, 
 as one can follow the non-ascending path to $y$ and then walk along the contour.

 So let $y$ be a point in the interface.

 However, since $y$ is in the interface, it is also wet.  Therefore for any contour $\phi$ which contains $y$, for any arbitrarily small $\eps$ there is 
 a wet point which lies above $\phi$ and has a descending path of length at most $\eps$ to $\phi$.
 This implies there is some contour which lies just 
 above $y$ which is wet.  So any contour $\phi$ containing $y$ has two contours which lie just above it, one which is dry, and one which is wet.  
 This can only happen if $\phi$ contains a merge vertex, because otherwise all contours which lie just above would be in the same L-homotopy equivalence class and so by 
 \Clm{mono}, if one is wet, all below it are wet.

 By similar logic one can argue that of the two contours through a given merge vertex at most one can contain interface points other than the merge vertex, 
 and that if one of the contours does contain such interface points, then the entire contour consists of interface points.
\end{proof}
}

@@

% \begin{proof} If $y \in \wet(x,\MM)$, but not in the interface, then any contour containing $y$ is also in $\wet(x,\MM)$.
% So consider a point $y$ in the interface. We argue that all points in a contour $\phi$ containing $y$
% are also in the interface. If such a contour is regular, we can find a contour just above it that
% is also wet. 
% 
% If two contours intersect, the intersection point $y$ is critical. We can argue that the only
% way to reach $y$ from a descending path is reach one of the contours and then follow it to get to $y$.
% But that would imply the existence of another critical point on this contour, violating the Morse condition.
% 
% Let us argue that $y$ is a merge vertex. Consider the descending path from $x$ to $y$. The final edge
% that reaches $y$ is obviously all wet, and is an increasing edge from $y$. Consider an opposite increasing
% edge. Take two point in these edges that are infinitesimally higher than $y$ and show that the contours
% through them as disjoint. So $y$ is a merge.
% \end{proof}

@@

\ignore{
As each original vertex in $\NN$ appears in only one of the sub-manifolds output by $\rain(x,\NN)$, we have the following corollary.

\begin{corollary}
The total size of a sub-manifolds output by $\rain(x,\MM)$ is $O(n)$.
\end{corollary}

\begin{corollary} \label{cor:rain-time} The total running time of $\rain(x,\MM)$ is $O(n)$.
\end{corollary}

\begin{proof} Consider some invocation to $\rain(x,\MM)$. The only non-trivial work done in $\rain$ is by the calls to $\wet$.
We know that each invocation to $\wet$ runs in time linear 
to the portion of the manifold which is wet (including the new boundary vertices created).  
As the output of $\rain$ is precisely the distinct sub-manifolds wet by each distinct call to $\wet$, the claim follows.
\end{proof}

%Consider one of the manifolds $\MM_i$ output by $\rain$.  This manifold has a set of boundary maxima/minima that are the result 
%of cutting along interface contours.  For each such contour there is exactly one other output manifold $\MM_j$ for which 
%the contour appears as a boundary maxima/minima. 
}

@@

\begin{proof} Consider the tree of recursive calls in $\rain(x,\MM)$, with each node
labeled with some $\MM_i$.
%
%There is a natural tree ordering defined over these manifolds where the 
%root corresponds the first manifold that was wet when we rained from $x$, and its children correspond to the recursive calls 
%(i.e. where we rained from the interface contours of the first raining).  Call this the \emph{Manifold Ordering Tree}.
%$\rain$ can easily be modified to also output this tree (without affecting the asymptotic running time).  
Walk through this tree in a leaf first ordering.  Each time we visit a node we connect its contour tree to the 
contour tree of its children in the tree using the $\surgery$ procedure. 
Each $\surgery$ call is constant time, and the total time is the size of the recursion tree.
%
%
%This takes $O(n)$ time overall by \Lem{new-verts} and since $\surgery$ takes constant time (given the appropriate vertices in each reeb graph).  The correctness follows by induction and  \Lem{surgery}.
\end{proof}

@@

%
%As part of the initialization, we determine
%all critical points in $\MM$. For any vertex $v$, we can look at $f(v) \pm \delta$-contours 
%in $f|_{B_\eps(v)}$ to determine the up and down degrees. This determines the criticality (and type)
%of $v$. This requires $O(|\MM|)$ time, since every facet is looked at proportional to the number of its
%vertices.

@@

\ignore{
Abusing notation, we say a contour $\phi$ is of color $c$ if all edges that intersect $\phi$ have color $c$.

\begin{definition} \label{def:color-full} Consider vertex $v$. An \emph{up-star} of $v$ is the set 
of edges that all connected to a fixed component of $\MM^+_v$. 
%and let $\phi$ be a $(f(v)+\delta)$-contour in
%$f|_{B_\eps(v)}$. If $\phi$ is monochromatic of color $c$, then $v$ is \emph{fully touched by $c$}. The set of colors
%that fully touch $v$ is denoted $\fcol(v)$.
\end{definition}

%Note that $\fcol(v)$ has size at most $2$, by the Morse condition. 
Note that $\col(v)$ 
%and $\fcol(v)$ can
be initially determined in $O(|\MM|)$ time, by a simple linear scan.
}

@@

%We initialize as follows. For each critical vertex $v$ and from each of its up-stars, we pick an arbitrary color $c$ touching $v$. 
%We insert $v$ into $T(c)$. Observe that $v$ participates in at least $1$ and at most $2$ heaps.
%

@@

\ignore{
\medskip
As the algorithm proceeds, it is convenient to imagine
it maintaining a set of contours $\Phi = \{\phi_1, \phi_2, \ldots, \phi_k\}$, which are contained
in incomparable edges of $\cC(\MM)$. Suppose we cut $\MM$ by each of these contours, and took
the simplicial complex $\NN$ where $\phi^-_i$'s are all boundary maxima. The algorithm will
have already determined the contour trees of the simplicial complexes
where $\phi^+_i$ is a boundary maxima.

During this process, various colors will be \emph{merged}. This involves merging the respective $T(c)$'s and
updating the union-find data structure. The only operation performed on $T(c)$'s are delete-max and merge.
No new vertex is ever inserted into the heaps. As colors merge, the initial painting is obviously modified.
But it would be too expensive to explicitly update colors. So the new painting is given by $\rep(c)$,
while we only have access to the original painting. By union-find bounds, this only adds an $\alpha(|\MM|)$
overhead to the data structure operations. The new coloring is guaranteed to be a painting of $\NN$.

%While this is the intuitive viewpoint, we describe how this information is actually maintained.
The vertices $\{\h(\rep(c))\}$ represent the set of contours $\Phi$. In terms of the contour tree, 
we have already constructed the contour trees rooted at each $h(\rep(c))$. This is trivially true at the initialization,
when each $h(c)$ is a maximum. One can now treat each $h(\rep(c))$ as a maximum, corresponding to (a possibly new) color $\rep(c)$.
To reduce confusion between the original painting and the new colors, we will assume that $\col(v)$ and $\fcol(v)$
are as given by the original painting. 
}

@@

\begin{lemma} \label{lem:full} Suppose vertex $v$ is connected to a component $\PP$ of $\MM^+_v$
by edge $e$ which is currently colored $c$. Either all edges in $\PP$ are currently colored $c$, or there
exists critical vertex $w \in \PP$ fully touched by $c$ and touched by another color.
%
%and take $(f(v)+\delta)$-contour $\phi$ intersecting
%$B_\eps(v)$. Suppose the edges incident to $v$ intersecting $\phi$ do not have the same color. For any such
%color $c$, there exists a critical vertex $w$ strictly higher than $v$ such that $c \in \fcol(v)$.
\end{lemma}

@@

\begin{lemma}
\label{lem:pathBounds} If $p$ is tall, $\sum_{v\in p} \log |H_v| = O(|p| \log |p|)$.
If $p$ is short, $\sum_{v\in p} \log |H_v| = O(|H_p| \log |H_p|)$.
% Let T be a binary tree and let $\chi(T)$ and $P(T)$ be a valid coloring and path decomposition of $T$, respectively.
% Then, 
% \begin{compactenum}
%  \item for any tall path $p\in P(T)$, $\sum_{v\in p} \log |H_v| = O(|p| \log |p|)$.
%  \item for any short path $p\in P(T)$, $\sum_{v\in p} \log |H_v| = O(|p| \log |H_{p}|)$.
% \end{compactenum}
\end{lemma}
\begin{proof} \ignore{
 By $\Obs{decrease}$ we know that for any vertex $v\in p$, $|H_v|\leq |H_{p}| + |p|$ (as $v$ is a descendant of $r_p$ along $p$).
}
For $v \in p$, $|H_v|\leq |H_{p}| + |p|$ (as $v$ is a descendant of $b(p)$ along $p$).
Hence, $\sum_{v\in p} \log |H_v| \leq \sum_{v\in p} \log(|H_{p}| + |p|)  = |p| \log (|H_{p}| + |p|)$.
 If $p$ is a tall path, then $|p| \log (|H_{p}| + |p|) = O(|p| \log |p|)$. If $p$ is short, then 
 $|p| \log (|H_{p}| + |p|) = O(|p| \log |H_{p}| )$. For short paths, $|p| = O(|H_p|)$.
\end{proof}

@@

%
%
%
%We focus on $\pmax(T)$, and prove the key property of this path decomposition.

@@

\section{Lower Bound by Path Decomposition}
\label{sec:lb}
We first prove a lower bound for join trees, and then generalize to contour trees.
%\Ben{Hopefully the following sentence clear up confusion between $\MM$ and $(f,\MM)$}
%Note earlier we usually used $\MM$ to denote the lift of a simplicial complex according to some height function $f$, 
%however in this section we make a distinction between the complex $\MM$ and the height function $f$.  
%In particular all previous definitions or subroutines which took $\MM$ as input now take $(f,\MM)$ as input.

@@

%In this section we prove \Thm{main-lb}, but for join trees instead of contour trees.  
%Observe that given a fixed $(f,\MM)$, 
%our algorithm for computing $\jc$ is deterministic and so the maximum path decomposition $P=P(\jc)$ defined by the initial coloring (\Def{initialColoring}) is fixed, 
%allowing us to define the cost, $\cost(f)=\sum_{p\in P} |p|\log|p|$.
%  
We focus on terrains, so $d=2$.
Consider any path decomposition $P$ of a valid join tree (i.e. any rooted binary tree).
When we say ``compute the join tree", we require the join tree to
be labeled with the corresponding vertices of the terrain.

%
%For given $\MM$, consider running the algorithm of \Sec{also} to compute $\jc(\MM)$.
%Let $P(\jc(\MM))$ be the maximum path decomposition of the join tree according to the initial coloring
%of $\init(\MM)$. Define $\cost_{\jc}(f) = \sum_{p\in P(\jc(\MM))} |p|\log|p|$

@@

% \begin{figure}[h]\centering
%     \includegraphics[width=.5\linewidth]{figs/bridge2}
%     \caption{On the left, the root of $P_i$ merges into a diamond of its parent $P_j$.  On the right, the root of $P_i$ splits off from a diamond of $P_j$}
%     \label{fig:diamond}
% \end{figure}

>>>>

Our algorithm requires numerous ideas: partitioning the contour tree into join and split trees, 
a local growing procedure to iteratively build contour trees, and the use of heavy path 
decompositions for the time complexity analysis. There is a crucial use of a family of binomial heaps to 
maintain priorities, ensuring that any comparison made is between nodes in the same descending path in the contour tree.
We also prove lower bounds showing that the $\sum_{p\in P} |p|\log|p|$ complexity
is inherent to computing contour trees. 
\end{abstract}

@@

In a contour tree $T$, a \emph{leaf path} is simply a contiguous path in $T$
containing a leaf, which is also monotonic in the height values of its vertices.

@@

We complement \Thm{main-alg} with a lower bound, suggesting that the comparison complexity
of $\sum_{p \in P(T)} |p|\log |p|$ is inherent for any contour tree algorithm. The exact phrasing
of the lower bound requires some care. Consider the $d=2$ (terrain) case.
Let a path decomposition $P$ be called \emph{valid} if it is actually output by the algorithm
on some input.

\begin{theorem} 
\label{thm:main-lb}
%Fix a sufficiently large numbers $t$ and $C$, 
%where $t$ is the number of critical points and $C$ is a valid cost for an instance with $t$ critical point.
Consider a valid path decomposition $P$. There exists a family $\bF_P$ of terrains ($d=2$) with the following properties.
Any contour tree algorithm makes $\Omega(\sum_{p \in P} |p|\log |p|)$ comparisons in the worst case over $\bF_P$.
Furthermore, for any terrain in $\bF_P$, the algorithm of \Thm{main-alg} makes $O(\sum_{p \in P} |p|\log |p|)$ comparisons.
%
%the set of functions $\bF_C$ that have the same number of critical points and the same (valid) cost $C$.  Any algebraic decision tree that correctly computes
%the contour tree on all of $\bF_C$ has a worst case running time of $\Omega(C)$.
\end{theorem}

@@

@@

@@

This is convenient, as we can now assume that $f$ is defined on $\RR^d$. Any point inside a boundary facet
has a well-defined height, and this includes the infinite facet. 
We think of $d$ as constant, and assume that $\MM$ is represented in a data structure that allow constant-time traversals of neighboring simplices
in $\MM$ (e.g.~\cite{BoMa12}). (This is analogous to a DCEL for higher dimensions.)
Observe that $f:\MM \rightarrow \RR$ can be thought of as a $d$-dimensional simplicial complex living in $\RR^{d+1}$.
We can think of $f(x)$ as the ``height" of a point $x \in \MM$, and encoded in the representation of $\MM$.

@@

@@

We use $\cV(f)$ to denote the set of critical points.
Because $f$ is piecewise-linear, all critical points are vertices in $\MM$. 
A value $h$ is called critical, if $f(v) = h$, for some $v \in \cV(f)$. 
The critical points are exactly where the topology of level sets change.
By assuming that our manifold is boundary critical, the vertices on a given boundary are either collectively all maxima or all minima.  
We abuse notation and refer to this entire set of vertices as a maxima or minima.

@@

@@

@@

An \emph{initial} painting has the additional property that the restriction $\chi:X \mapsto [|X|]$ is a bijection.
\end{definition}

@@

@@

@@

@@

@@

\begin{proof} Consider the tree of recursive calls in $\rain(x,\MM)$, with each node
labeled with some $\MM_i$.
Walk through this tree in a leaf first ordering.  Each time we visit a node we connect its contour tree to the 
contour tree of its children in the tree using the $\surgery$ procedure. 
Each $\surgery$ call is constant time, and the total time is the size of the recursion tree.
\end{proof}

@@

@@

@@

@@

@@

\begin{lemma} \label{lem:full} Suppose vertex $v$ is connected to a component $\PP$ of $\MM^+_v$
by edge $e$ which is currently colored $c$. Either all edges in $\PP$ are currently colored $c$, or there
exists critical vertex $w \in \PP$ fully touched by $c$ and touched by another color.
\end{lemma}

@@

\begin{lemma}
\label{lem:pathBounds} If $p$ is tall, $\sum_{v\in p} \log |H_v| = O(|p| \log |p|)$.
If $p$ is short, $\sum_{v\in p} \log |H_v| = O(|H_p| \log |H_p|)$.
\end{lemma}
\begin{proof} \ignore{
 By $\Obs{decrease}$ we know that for any vertex $v\in p$, $|H_v|\leq |H_{p}| + |p|$ (as $v$ is a descendant of $r_p$ along $p$).
}
For $v \in p$, $|H_v|\leq |H_{p}| + |p|$ (as $v$ is a descendant of $b(p)$ along $p$).
Hence, $\sum_{v\in p} \log |H_v| \leq \sum_{v\in p} \log(|H_{p}| + |p|)  = |p| \log (|H_{p}| + |p|)$.
 If $p$ is a tall path, then $|p| \log (|H_{p}| + |p|) = O(|p| \log |p|)$. If $p$ is short, then 
 $|p| \log (|H_{p}| + |p|) = O(|p| \log |H_{p}| )$. For short paths, $|p| = O(|H_p|)$.
\end{proof}

@@

@@

\section{Lower Bound by Path Decomposition}
\label{sec:lb}
We first prove a lower bound for join trees, and then generalize to contour trees.

@@

We focus on terrains, so $d=2$.
Consider any path decomposition $P$ of a valid join tree (i.e. any rooted binary tree).
When we say ``compute the join tree", we require the join tree to
be labeled with the corresponding vertices of the terrain.

@@
